Anomalous waves triggered by abrupt depth changes: experiments and truncated KdV theory

### 2019-10-10 paper2_JNLS/paper.tex
NM

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
%\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation & Experimental range & `Representative' values \\
\hline
Peak forcing frequency	& $f_p$					& 2 Hz 		& 2 Hz \\
Characteristic amplitude	& $\etastd$				& 0.03--0.3 cm	& 0.21 cm \\
Upstream depth		& $\dup$					& 12.5 cm 	& 12.5 cm \\
Downstream depth		& $\ddn$					& 2.2--5.3 cm	& 3 cm \\
Upstream wavelength	& $\lamup = \sqrt{g \dup}/f_p$	& 55 cm 		& 55 cm \\
Downstream wavelength	& $\lamdn = \sqrt{g \ddn}/f_p$	& 23--36 cm	& 27 cm \\
%
Amplitude-to-depth ratio	& $\epsup = \etastd / \dup$	&0.0024--0.024	& 0.017 \\
Depth-to-wavelength ratio	& $\delup = \dup / \lamup$	& 0.23		& 0.23 \\
Depth ratio			& $\dratdn = \ddn/\dup$		& 0.18--0.42	& 0.24 \\
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

\subsection{Calibration of the inverse temperature}

@@

\subsection{Statistical comparison between theory and experiments}

@@

	The comparison between theory and experiment is striking. As seen in the figure, both the experiments and theory show a transition from a nearly symmetric upstream distribution, to a highly skewed distribution downstream. The TKdV theory not only captures this transition, but also recovers the shape of the resulting downstream distribution in remarkable detail. Nearly every feature that can be compared --- the decay rate of the tail, the position and value of the peak, the rapid cutoff for negative $u$ --- matches surprisingly well. This comparison offers compelling visual evidence for: (a) the predictive power of the TKdV framework, and (b) the successful calibration of the incoming inverse temperature.

@@

\subsection{Wave dynamics}

Next, to complement the statistical comparison, we investigate some individual solution trajectories from the TKdV simulations.

\vsp{5}
We show individual trajectories in the next figure for upstream and downstream. In the upstream flow, due to the coefficients, the nonlinearity is very weak. Hence, wave propagation is dominated by linear dispersion. Indeed, with weak nonlinearity, \eqref{dimlessKdV} is approximated by
\begin{equation}
u_t + C_2 \drat^{1/2} \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
\end{equation}
which implies the dispersion relation $\omega_k = -C_2 k^3$ or for phase velocity $c_k = -C_2 k^2$, which implies left-going waves \cite{majdaqi2019}.
% Further details, see page 6 of Majda Qi, J Stat Phys.

@@

>>>>

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
%\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation & Experimental range & `Representative' values \\
\hline
Peak forcing frequency	& $f_p$					& 2 Hz 		& 2 Hz \\
Characteristic amplitude	& $\etastd$				& 0.03--0.3 cm	& 0.21 cm \\
Upstream depth		& $\dup$					& 12.5 cm 	& 12.5 cm \\
Downstream depth		& $\ddn$					& 2.2--5.3 cm	& 3 cm \\
Upstream wavelength	& $\lamup = \sqrt{g \dup}/f_p$	& 55 cm 		& 55 cm \\
Downstream wavelength	& $\lamdn = \sqrt{g \ddn}/f_p$	& 23--36 cm	& 27 cm \\
%
Amplitude-to-depth ratio	& $\epsup = \etastd / \dup$	&0.0024--0.024	& 0.017 \\
Depth-to-wavelength ratio	& $\delup = \dup / \lamup$	& 0.23		& 0.23 \\
Depth ratio			& $\dratdn = \ddn/\dup$		& 0.18--0.42	& 0.24
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%----------------------------------------------------------------------------%
% Calibration of inverse temperature
%----------------------------------------------------------------------------%
\subsection{Calibration of the inverse temperature}

@@

%----------------------------------------------------------------------------%
% Comparison of distributions
%----------------------------------------------------------------------------%
\subsection{Statistical comparison between theory and experiments}

@@

	The comparison between theory and experiment in Fig.~\ref{uhist} is striking. As seen in the figure, both the experiments and theory show a transition from a nearly symmetric upstream distribution, to a highly skewed distribution downstream. The TKdV theory not only captures this transition, but also recovers the shape of the resulting downstream distribution in remarkable detail. Nearly every feature that can be compared --- the decay rate of the tail, the position and value of the peak, the rapid cutoff for negative $u$ --- matches surprisingly well. This comparison offers compelling visual evidence for: (a) the predictive power of the TKdV framework, and (b) the successful calibration of the incoming inverse temperature.

@@

%----------------------------------------------------------------------------%
% Wave dynamics
%----------------------------------------------------------------------------%
\subsection{Wave dynamics}

	As a complement to the above statistical comparison, we now examine the wave dynamics of a few individual trajectories from the TKdV simulations. Figure \ref{trajectories} shows example upstream and downstream solution trajectories from the same TKdV simulations that were used to produce the histograms in Figs.~\ref{uhist}--\ref{slopehist}. The dimensionless displacement, $u$, is represented by color in the domain $(x,t) \in [-\pi,\pi] \times [-10,10]$, where the ADC is encountered at $t=0$.

	Visual differences between the upstream and downstream dynamics are apparent in Fig.~\ref{trajectories}. The upstream trajectory shows several waves of modest amplitude all propagating leftward. Here, the magnitude of the positive and negative displacements are comparable. This is in stark contrast to the downstream dynamics, which feature fewer waves of larger amplitude propagating in either direction. In particular, one large wave is seen to propagate from the lower-left corner of the figure to the upper right. A bias towards positive surface displacement is visually apparent in these downstream dynamics.

@@

These observations can be rationalized with some simple analysis of the structure of TKdV. Upstream of the ADC, since $\drat = 1$, the nonlinear effects are relatively weak and the dynamics are dominated by dispersion. Linearizing Eq.~\eqref{TKdV} and introducing the ansatz $u_k = e^{i(kx - \omega_k t)}$ produces the dispersion relation for angular frequency $\omega_k = -C_2 k^3$, or for phase velocity $c_k = \omega_k/k = -C_2 k^2$ \cite{majdaqi2019}. In particular, the phase velocity is strictly negative, implying that waves can only propagate leftwards, as is consistent with Fig.~\ref{trajectories}(a).
% Further details, see page 6 of Majda Qi, J Stat Phys.
%$C_3 = 0.20$ and $C_2 = 0.024$.

Downstream of the ADC, however, $\drat = 0.24$, which significantly amplifies nonlinearity while suppressing dispersion in Eq.~\ref{TKdV} and allows waves to propagate in either direction. In particular, the dispersionless limit of KdV is the Burgers-Hopf equation, in which the nonlinear advection is proportional in both amplitude and direction to the value of $u$. Hence, positive displacements would be expected to move rightwards, as is seen in Fig.~\ref{trajectories}(b).

%We remind the reader that the KdV framework tracks the long-time evolution of waves in the reference frame moving with the linear speed $c = \sqrt{g \depth}$.

% Timescale of 10 is about right to see waves propagate through the entire periodic domain. This however would be a very long experimental time scale. So what gives? We answer that question with skewness and kurtosis plots which evolve over a much shorter timescale.

\vsp{5}

Next stuff
